---
title: "Introduction to ClassTopics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ClassTopics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Stan fitting requires a working cmdstan installation;
                # code is shown but not executed when this vignette is built.
)
```

## Overview

`ClassTopics` fits a **supervised topic model** to count data (for example,
gene expression counts) together with a binary or categorical response. It
combines two components, estimated jointly:

1. A topic model that integrates non-negative matrix factorization (NMF)
   and latent Dirichlet allocation (LDA) to factor the count matrix into
   patient-level topic proportions (`theta`) and topic-level variable
   profiles (`beta`).
2. A multinomial/categorical regression of the response on those topic
   proportions (coefficients `eta`), fit at the same time as the topics
   themselves, so the discovered topics are shaped by their ability to
   predict the outcome.

Estimation is done with full Bayesian inference in
[Stan](https://mc-stan.org/), via
[`cmdstanr`](https://mc-stan.org/cmdstanr/). An EM warm-start initializes
the sampler for faster, more stable convergence.

This vignette walks through:

- fitting a model with `ClassTopics()`
- interpreting the result with `ClassTopics_results()`
- diagnostic plots
- cross-validation with `cv_ClassTopics()`
- predicting on new samples

## Setup

```{r setup}
library(ClassTopics)
```

`ClassTopics` requires a working `cmdstanr` installation. If you haven't
set this up yet:

```{r cmdstan-setup}
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
cmdstanr::install_cmdstan()
```

This is a one-time setup per machine. The Stan models bundled with
`ClassTopics` are compiled automatically the first time you call a fitting
function, and the compiled executables are cached and reused afterwards.

## Simulating example data

To keep this vignette self-contained, we simulate a small gene-count
dataset with a categorical response driven by three latent topics.

```{r simulate-data}
set.seed(123)

D <- 120   # number of patients
V <- 60    # number of genes
K <- 6     # true number of topics
C <- 3     # number of response categories

# Simulate topic-gene profiles (beta) and patient topic proportions (theta)
true_beta  <- matrix(rgamma(K * V, shape = 2, rate = 2), nrow = K, ncol = V)
true_theta <- matrix(rgamma(D * K, shape = 2, rate = 2), nrow = D, ncol = K)
true_theta <- true_theta / rowSums(true_theta)

# Simulate gene counts: Poisson NMF
lambda <- true_theta %*% true_beta * 200
gene_counts <- matrix(
  rpois(D * V, lambda = as.vector(lambda)),
  nrow = D, ncol = V
)
colnames(gene_counts) <- paste0("gene_", seq_len(V))

# Simulate a response driven by the topic proportions
true_eta <- matrix(rnorm(C * K, sd = 2), nrow = C, ncol = K)
linear_pred <- true_theta %*% t(true_eta)
probs <- exp(linear_pred) / rowSums(exp(linear_pred))
response <- factor(
  apply(probs, 1, function(p) sample(seq_len(C), size = 1, prob = p)),
  labels = c("A", "B", "C")
)

table(response)
```

## Fitting a model

```{r fit}
fit <- ClassTopics(
  counts        = gene_counts,
  response      = response,
  K             = 3,
  betadir       = TRUE,
  seed          = 123,
  chains        = 3,
  cores         = 3,
  iter_warmup   = 1000,
  iter_sampling = 1000
)
```

`ClassTopics()` returns a compiled and fitted `cmdstanr` `CmdStanMCMC`
object. Use `ClassTopics_results()` to turn this into an interpretable,
structured `CTresults` object.

## Interpreting the results

```{r results}
results <- ClassTopics_results(
  fit               = fit,
  true_response     = response,
  vars_names        = colnames(gene_counts),
  top_vars          = 10,
  credible_interval = 0.95
)

results
```

`results` is an S4 `CTresults` object. You can pull out specific pieces
with its accessors:

```{r accessors}
getPredictions(results)          # predicted class + probabilities
regressionCoefficients(results)  # eta estimates, with credible intervals
topVars(results)                # top genes per topic
```

## Diagnostic plots

```{r plots}
plot_topic_correlations(results)
plot_topic_response_heatmap(results)
plot_top_vars(results, n_vars = 5)
```

`plot_topic_correlations()` shows how correlated the estimated topics are
with one another -- useful for spotting redundant topics.
`plot_topic_response_heatmap()` shows which topics are most strongly
associated with which response categories. `plot_top_vars()` shows the
highest-probability vars driving each topic.

## Cross-validation

A single fit's in-sample accuracy will overstate real-world performance.
`cv_ClassTopics()` runs stratified k-fold cross-validation, fitting one
model per fold (plus one final model on the full data), in parallel via
the `future` package:

```{r cv}
cv_fit <- cv_ClassTopics(
  counts        = gene_counts,
  response      = response,
  k_folds       = 5,
  K_topics      = 3,
  betadir       = TRUE,
  seed          = 123,
  iter_warmup   = 1000,
  iter_sampling = 1000
)

cv_fit
```

`cv_ClassTopics()` fits one model per fold plus one final model on the
full dataset (run concurrently via the `future` package), and returns a
`cvCTprediction_plusfm` object with the pooled cross-validated diagnostics
and the final full-data fit. Key accessors:

```{r cv-accessors}
cvTestAccuracy(cv_fit)   # pooled out-of-fold test accuracy
overfittingGap(cv_fit)   # train accuracy - test accuracy
```

A large overfitting gap suggests the model (or `K_topics`, or `sigma_eta`)
may need adjusting -- for example, increasing `sigma_eta` or `lambda_ridge`
to regularize `eta` more strongly.

The final model, fit on the complete dataset, is available via
`cv_fit@final_model` -- this is the same kind of `cmdstanr` fit object
returned by `ClassTopics()`, and can be summarized the same way:

```{r cv-final-model}
cv_results <- ClassTopics_results(
  fit = cv_fit@final_model,
  true_response = response,
  vars_names = colnames(gene_counts)
)
```

## Predicting on new patients

Given a fitted model's learned topic-variable profiles (`beta`) and regression
coefficients (`eta`), you can predict the response for new observations from
their counts alone, without knowing their true response.

Two prediction routes are available:

- `predict_ClassTopics_stan()`: refits topic proportions for the new
  patients via Stan, fully Bayesian (matches how the original model was fit).
- `predict_ClassTopics_EM()`: a faster, EM-only approximation, useful when
  Stan's overhead isn't justified (e.g. very large test sets, or quick
  exploratory checks).

```{r predict-stan}
# results@beta and results@eta are CTparameter objects (posterior mean +
# credible interval); extract the plain posterior-mean matrices with
# paramMean() for use as plug-in estimates in prediction.
W   <- paramMean(results@beta)
eta <- paramMean(results@eta)

test_pred <- predict_ClassTopics_stan(
  counts_test = gene_counts_new,  # a D_test x V matrix, same variables/order
  W = W,
  eta = eta,
  mu_target = sqrt(mean(gene_counts) / nrow(W)),
  response = response,   # original training response factor (its levels
                         # define the prediction categories)
  seed = 123
)

test_pred$predicted_class
test_pred$predicted_probs
```

```{r predict-em}
test_pred_em <- predict_ClassTopics_EM(
  counts_test = gene_counts_new,
  W = W,
  eta = eta,
  mu_target = sqrt(mean(gene_counts) / nrow(W)),
  response = response
)

test_pred_em$predicted_class
```

## Summary

- `ClassTopics()` fits the joint NMF + supervised topic model via Stan.
- `ClassTopics_results()` turns a raw fit into an interpretable `CTresults`
  object, with accessors for predictions, coefficients, and top variables.
- `plot_topic_correlations()`, `plot_topic_response_heatmap()`, and
  `plot_top_vars()` provide diagnostic visualizations.
- `cv_ClassTopics()` provides an honest, cross-validated estimate of
  predictive performance and an overfitting diagnostic.
- `predict_ClassTopics_stan()` / `predict_ClassTopics_EM()` predict the
  response for new, unlabeled samples from a fitted model.
