---
title: "Introduction to ClassTopics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ClassTopics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Stan fitting requires a working cmdstan installation;
                # code is shown but not executed when this vignette is built.
)
```

## Overview

`ClassTopics` fits a **supervised topic model** to count data (for example,
gene expression counts) together with a binary or categorical response. It
combines two components, estimated jointly:

1. A topic model that integrates non-negative matrix factorization (NMF)
   and latent Dirichlet allocation (LDA) to factor the count matrix into
   patient-level topic proportions (`theta`) and topic-level variable
   profiles (`beta`).
2. A multinomial/categorical regression of the response on those topic
   proportions (coefficients `eta`), fit at the same time as the topics
   themselves, so the discovered topics are shaped by their ability to
   predict the outcome.

Estimation is done with full Bayesian inference in
[Stan](https://mc-stan.org/), via
[`rstan`](https://mc-stan.org/rstan/). An EM warm-start initializes
the sampler for faster, more stable convergence.

This vignette walks through:

- generating a count dataset with `generate_CTdata()`
- fitting a model with `ClassTopics()`
- interpreting the result with `ClassTopics_results()`
- diagnostic plots
- cross-validation with `cv_ClassTopics()`
- predicting on new samples

## Setup

```{r setup}
library(ClassTopics)
```

## Generating example data

To keep this vignette self-contained, we simulate, for instance, a small gene-count
dataset with a categorical response driven by three latent topics.

## Generating count data

To keep this vignette self-contained, we generate, for instance, a small count
dataset with a 3-level categorical response driven by five latent topics using
`generate_CTdata()`.

```{r simulate-data}
set.seed(123)

C <- 3                         # Number of classes
eta <- matrix(rnorm(15), C, 5) # Topic-response matrix

D_per_class <- 50              # Number of observations per class
V <- 60                        # Number of variables
alpha_w <- 0.05                # Controls data sparsity
lambda <- 1e7                  # Magnitude of the dataset
class_spread <- 0.6            # Controls levels of class separation
seed <- 123

# Data simulation
df <- generate_CTdata(
  D_per_class = D_per_class,
  V = V,
  eta = eta,
  alpha_w = alpha_w,
  lambda = lambda, 
  class_spread = class_spread,
  seed = seed)

tpc <- 10                      # Number of test observations per class

test_ids <-
  rep(D_per_class * 0:(C-1), each = tpc) +
  rep((D_per_class-tpc+1):D_per_class, C)

gene_counts_train <- df@X[-test_ids, ]
response_train <- df@Y[-test_ids]

gene_counts_test <- df@X[test_ids, ]
response_test <- df@Y[test_ids]
```

## Fitting a model

```{r fit}
fit <- ClassTopics(
  counts        = gene_counts_train,
  response      = response_train,
  K             = 3,
  betadir       = TRUE,
  seed          = 123,
  chains        = 3,
  cores         = 3,
  iter_warmup   = 1000,
  iter_sampling = 1000
)
```

`ClassTopics()` returns a compiled and fitted `rstan`
object. Use `ClassTopics_results()` to turn this into an interpretable,
structured `CTresults` object.

## Interpreting the results

```{r results}
results <- ClassTopics_results(
  fit               = fit,
  true_response     = response_train,
  vars_names        = colnames(gene_counts_train),
  top_vars          = 10,
  credible_interval = 0.95
)

results
```

`results` is an S4 `CTresults` object. You can pull out specific pieces
with its accessors:

```{r accessors}
getPredictions(results)          # predicted class + probabilities
regressionCoefficients(results)  # eta estimates, with credible intervals
topVars(results)                # top genes per topic
```

## Diagnostic plots

```{r plots}
plot_topic_correlations(results)
plot_topic_response_heatmap(results)
plot_top_vars(results, n_vars = 5)
```

`plot_topic_correlations()` shows how correlated the estimated topics are
with one another -- useful for spotting redundant topics.
`plot_topic_response_heatmap()` shows which topics are most strongly
associated with which response categories. `plot_top_vars()` shows the
highest-probability vars driving each topic.

## Cross-validation

A single fit's in-sample accuracy will overstate real-world performance.
`cv_ClassTopics()` runs stratified k-fold cross-validation, fitting one
model per fold (plus one final model on the full data), in parallel via
the `future` package:

```{r cv}
cv_fit <- cv_ClassTopics(
  counts        = gene_counts_train,
  response      = response_train,
  k_folds       = 5,
  K_topics      = 3,
  betadir       = TRUE,
  seed          = 123,
  iter_warmup   = 1000,
  iter_sampling = 1000
)

cv_fit
```

`cv_ClassTopics()` fits one model per fold plus one final model on the
full dataset (run concurrently via the `future` package), and returns a
`cvCTprediction_plusfm` object with the pooled cross-validated diagnostics
and the final full-data fit. Key accessors:

```{r cv-accessors}
cvTestAccuracy(cv_fit)   # pooled out-of-fold test accuracy
overfittingGap(cv_fit)   # train accuracy - test accuracy
```

A large overfitting gap suggests the model (or `K_topics`, or `sigma_eta`)
may need adjusting -- for example, increasing `sigma_eta` or `lambda_ridge`
to regularize `eta` more strongly.

The final model, fit on the complete dataset, is available via
`cv_fit@final_model` -- this is the same kind of `cmdstanr` fit object
returned by `ClassTopics()`, and can be summarized the same way:

```{r cv-final-model}
cv_results <- ClassTopics_results(
  fit = cv_fit@final_model,
  true_response = response_train,
  vars_names = colnames(gene_counts)
)
```

## Predicting on new patients

Given a fitted model's learned topic-variable profiles (`beta`) and regression
coefficients (`eta`), you can predict the response for new observations from
their counts alone, without knowing their true response.

Two prediction routes are available:

- `predict_ClassTopics_stan()`: refits topic proportions for the new
  patients via Stan, fully Bayesian (matches how the original model was fit).
- `predict_ClassTopics_EM()`: a faster, EM-only approximation, useful when
  Stan's overhead isn't justified (e.g. very large test sets, or quick
  exploratory checks).

```{r predict-stan}
# results@beta and results@eta are CTparameter objects (posterior mean +
# credible interval); extract the plain posterior-mean matrices with
# paramMean() for use as plug-in estimates in prediction.
W   <- paramMean(results@beta)
eta <- paramMean(results@eta)

test_pred <- predict_ClassTopics_stan(
  counts_test = gene_counts_test, 
  W = W,
  eta = eta,
  response_test = response_test,
  seed = 123
)

test_pred@predicted_class
test_pred@predicted_probs
```

```{r predict-em}
test_pred_em <- predict_ClassTopics_EM(
  counts_test = gene_counts_test,
  W = W,
  eta = eta,
  response_test = response_test
)

test_pred_em@predicted_class
```

## Summary

- `generate_CTdata()` generates synthetic count datasets.
- `ClassTopics()` fits the joint NMF + supervised topic model via Stan.
- `ClassTopics_results()` turns a raw fit into an interpretable `CTresults`
  object, with accessors for predictions, coefficients, and top variables.
- `plot_topic_correlations()`, `plot_topic_response_heatmap()`, and
  `plot_top_vars()` provide diagnostic visualizations.
- `cv_ClassTopics()` provides an honest, cross-validated estimate of
  predictive performance and an overfitting diagnostic.
- `predict_ClassTopics_stan()` / `predict_ClassTopics_EM()` predict the
  response for new, unlabeled samples from a fitted model.
